---
title: "Scientific Computing for Biologists - Assignment 1"
author: "Carson Stacy"
date: "2022-09-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Duck, Duck, Goose?

## An anatidaephobic's guide to the state of New York

Some people are afraid of ducks and geese, and I don't blame them. They are horrifying little honkers. Thankfully, with the help of the crowdsourced Ebird dataset maintained by Cornell University, we can learn more about where these menaces are most often seen so they can be properly avoided.

```{r loading our data, echo=F, include=F}
library(stringr)
library(tidyverse)
library(here)
library(lubridate)
library(ggmap)
library(sf)
library(osmdata)
library(showtext)
library(ggsci)
library(magrittr)

# read in the data
tidyDuckGoose <- readr::read_delim(here("tidyDuckGoose.tsv"),
                          delim = "\t", 
                          escape_double = FALSE, 
                          col_names = FALSE, 
                          trim_ws = TRUE)
```

```{r echo=F, include=F}
# Data processing
tidyDuckGoose_filt <- tidyDuckGoose %>%
  dplyr::rename("bird_type" = X1, 
                "name" = X2, 
                "species" = X3, 
                "lat" = X4, "lon" = X5, 
                "date" = X6)  %>% # add column names
  dplyr::mutate(date = lubridate::as_date(date)) %>% # fix date formatting
  drop_na() %>% # drop naughty data (n=5)
  filter(name %in% c("American Black Duck", "Barnacle Goose", "Brant",
                     "Cackling Goose", "Canada Goose", "Long-tailed Duck", 
                     "Ring-necked Duck", "Ruddy Duck", "Wood Duck")) 
                      # getting rid of undesired species groups (n=25)

```


## You're not crazy, it is getting worse!

As you can see in figure 1 below, the number of spottings of these birds are on the rise. If this pattern continues, we will soon be inundated with human-goose conflict. 

```{r timeseries, echo=FALSE, include=T}
MyPalette <- c("Canada Goose" = "#8491B4FF", "Brant" = "#e64B35FF", "Cackling Goose" = "#F39B7FFF", "Barnacle Goose" = "#DC0000FF",
               "American Black Duck" = "#3C5488FF", "Long-tailed Duck" = "#4DBBD5FF", "Ring-necked Duck" = "#00A087FF", "Ruddy Duck" =  "#91D1C2FF", "Wood Duck" = "#7E6148FF"
               )

tidyDuckGoose_filt %>%
  group_by(name, year(date)) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = `year(date)`, y = count, group = name, color = name)) + 
  geom_line(size=1.5) +
  theme_classic() +
  scale_colour_manual(values = MyPalette) +
  theme(legend.position = "bottom", legend.title= element_blank()) + 
  xlab("Year") +
  ylab("Number of individuals identified")
  # scale_color_npg()
```


```{r map prep, echo=FALSE, include=F}
NY_bb <- osmdata::getbb(place_name = "New York State")

# NY_bb

map <- get_stamenmap(bbox = NY_bb, zoom = 8, maptype = "toner-lite")
ggmap(map)

ny <- map_data("state", region="new york")
```


We can compare the geographical distributions of ducks and geese, note the overlap in distributions
```{r map, echo=FALSE, include=T}
ggmap(map) + 
  geom_polygon(data = ny, aes(x=long, y=lat), color = "darkgrey", alpha = 0) +
  geom_count(data=tidyDuckGoose_filt,
             aes(x=signif(lon,6) ,y=signif(lat,6),color=name),
             # size=1,
             alpha=.5) +
  facet_wrap(~fct_rev(bird_type)) +
  theme_classic() +
  scale_colour_manual(values = MyPalette) +
  # scale_color_npg() +
  labs(caption = "Data: Ebird Archive | Plot: @clstacy",
       title = 'Fascinating fowel and where to find them',#) +#"\U1FABF") + #U+1F986
       x = 'longitude',
       y = 'latitude') +
  theme(legend.position = "bottom", legend.title= element_blank())
```


It appears there are regions, including near NYC, where many of these birds are seen.
```{r map2, echo=FALSE, include=T}
ggmap(map) + 
  geom_polygon(data = ny, aes(x=long, y=lat), color = "darkgrey", alpha = 0) +
  geom_count(data=tidyDuckGoose_filt,
             aes(x=signif(lon,6) ,y=signif(lat,6),color=name),
             # size=1,
             alpha=.5) + 
  theme_classic() +
  # scale_color_npg() +
  scale_colour_manual(values = MyPalette) +
  labs(caption = "Data: Ebird Archive | Plot: @clstacy",
       title = 'Fascinating fowel and where to find them',#) +#"\U1FABF") + #U+1F986
       x = 'longitude',
       y = 'latitude') +
  theme(legend.position = "bottom", legend.title= element_blank())

# 
# ggplot() +
#   geom_sf(data = bass_strait_polygon, fill = "#33658A", colour = NA) +
#   geom_sf(data = yarra_line, colour = "#33658A", size = 1.5) +
#   geom_sf(data = melbourne_trainline_lines, size = 1.2, colour = "gray40") +
#   #Add a label for port phillip bay and bass strait
#   geom_text(aes(label = "Port Phillip Bay"), x = 144.859280, y = -38.092014,
#             family = "Alice", fontface = "bold", colour = "grey90") +
#   geom_text(aes(label = "Bass Strait"), x = 144.513047, y = -38.421423,
#             family = "Alice", fontface = "bold", colour = "grey90") +
#   #Add title as text inside the plot
#   geom_text(aes(label = "- Trains of Melbourne -"), x = 145.75, y = -37.43,
#             family = "Alice", size = 8,
#             colour = "grey45", fontface = "bold") +
#   geom_text(aes(label = "Using OpenStreetMap data"), x = 145.75, y = -37.5,
#             family = "Alice", size = 4,
#             colour = "grey45", fontface = "bold") +
#   labs(caption = "Data: OSM | Plot: @ldbailey255") +
#   coord_sf(xlim = melb_bb[1, ], ylim = melb_bb[2, ]) +
#   theme_classic() +
#   theme(panel.background = element_rect(fill = "#f2eadf"),
#         panel.border = element_rect(fill = NA, colour = "grey45", size = 1.5),
#         axis.text = element_text(family = "Alice", size = 14),
#         plot.caption = element_text(family = "Ubuntu Mono", size = 12,
#                                   colour = "grey45"),
#         axis.line = element_blank())
```


```{r}
ggplot(tidyDuckGoose_filt, aes(x=species, fill = name)) + 
  geom_bar() + 
  theme_classic() +
  # scale_color_npg() +
  scale_fill_manual(values = MyPalette) +
  coord_flip() +
  facet_wrap(~bird_type, scales = "free") +
  geom_text(stat='count', aes(label=..count..), vjust=0.5, hjust=0.5, angle = -90) +
  theme(legend.title= element_blank()) +
  ggtitle("Number of each species identified in the data")
             
```

```{r map_months, echo=FALSE, include=T}
tidyDuckGoose_filt %>%
  mutate(month = lubridate::month(date, label = T)) %>%
  group_by(name, month) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = month, y = count, group = name, color = name)) + 
  geom_line(size=1.5) +
  theme_classic() +
  scale_colour_manual(values = MyPalette) +
  theme(legend.position = "bottom", legend.title= element_blank()) + 
  xlab("Month") +
  ylab("Seasonal trends in identification")
```
As we can see, geese are omnipresent. However, July does seem to be the best time for avoiding seeing ducks and geese in general.

Broken down by year, we can see that this trend is conserved, but sightings are rapdily increasing.
```{r map_months_yrs, echo=FALSE, include=T}
tidyDuckGoose_filt %>%
  mutate(month = lubridate::month(date, label = T),
         year = lubridate::year(date)) %>%
  filter(year > 2004) %>%
  group_by(name, month, year) %>%
  summarise(count = n()) %>%
  ggplot(aes(x = month, y = count, group = name, color = name)) + 
  geom_line() +
  geom_point(size = 0.5) +
  theme_classic() +
  scale_colour_manual(values = MyPalette) +
  theme(legend.position = "bottom", legend.title= element_blank()) + 
  xlab("Month") +
  ylab("Seasonal trends in identification") + 
  facet_wrap(~ year) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```
